Bridging Gaps: Investigating COVID-19's Influence on Health Disparities in Connecticut

### 2023-12-15 manuscript/manuscript.tex
added analysis of education levels to discussion

\section{Introduction}\label{sec:intro}
\\
I. Prevalence of Social Determinants of Health 
\\
Current research focuses on how social determinants of health (SDoH) plays a  massive
impact on one's health; it is estimated that 80 percent of a population's health outcomes are 
dictated by SDoH \citep{HOOD2016129}. Often, SDoH, when referring to an individual, can result in racial 
disparities in care when looking at a population\citep{Monroe2023-uq}. It has been shown that major inefficiencies
in the health system are attributed to overlooked prevention opportunities and unequal access
to care.\citep{Allin2014-xn}

@@

Despite the growing body of research on SDoH, there is a notable gap concerning the specific impact of COVID-19 on 
these determinants within diverse communities. This research aims to bridge this gap by comprehensively assessing how 
the pandemic has influenced key social determinants of health across the 8 CT counties and a four year timeframe prior to 
the start of the COVID pandemic. 
By identifying the specific ways in which different communities were affected, this study contributes valuable insights 
for targeted interventions, policy-making, and the development of equitable healthcare strategies.
\\
II. Introduction to the Impact of Social Needs(specify for CT)
\\
This study will compare each of the following social needs by Year(2017-2020) and by 8 CT Counties: 
Median Income,
Educaation,
Food Access
Housing Stability,
Health Insurance,
Reshospitalization rates.

Income is a fundamental
social need that significantly influences
one's health. The level of income
directly affects an individual's access
to essential resources such as nutritious
food, healthcare, and secure housing.
People with lower incomes often face
barriers to healthcare services and
struggle to afford a balanced diet,
leading to higher risks of chronic
illnesses. Income disparities contribute
to inequalities in health outcomes, as
those with limited financial resources
may experience higher stress levels and
reduced access to preventive care.
Addressing income inequality is crucial
for promoting overall well-being and
reducing health disparities in communities.

Education is a key social determinant
of health, influencing various aspects
of an individual's life. Higher levels
of education are associated with better
health outcomes, as education enhances
one's ability to make informed decisions
about lifestyle choices, healthcare
utilization, and disease prevention.
Limited educational opportunities can
result in lower health literacy, reducing
an individual's capacity to navigate
complex healthcare systems and understand
preventive measures. Educational
disparities contribute to a cycle of
poor health outcomes, as individuals
with lower levels of education may face
challenges in securing stable employment
and accessing adequate healthcare
resources. Addressing educational
inequalities is essential for fostering
a healthier and more equitable society.

Access to nutritious and sufficient
food is a critical social need that
directly impacts health. Food insecurity,
often linked to low income, can lead to
malnutrition, obesity, and various health
issues. Lack of access to fresh and
healthy foods contributes to the prevalence
of chronic conditions such as diabetes
and cardiovascular diseases. Current
issues related to food include disparities
in food availability, affordability, and
nutritional quality, disproportionately
affecting marginalized communities.
Addressing these issues involves promoting
food equity, supporting local agriculture,
and implementing policies that ensure
everyone has access to nutritious food,
ultimately improving overall public health.

@@

\ref{fig:Education by County Graph}
discuss, use lit to support why(descriptive) specific to CT

Table 4\ref{fig:Social Needs by County Graph} reveals that there is a signifiant difference between 
social need variables by year (2017-2020) for median income, food, and uninsured rate. Table 4 also 
reveals that all five variables analyzes(median income, food, housing, rehospitalization rate, and 
uninsured rate) are significantly difference. Multiple Comparison tests are conducted to identify the 
specific similarities and differences between the 8 CT counties.
\\
I. Comparison by County
\\
The results of the Tukey multiple comparison analysis\ref{tab: Multiple Comparisons} shed light on the relationship between county and key 
variables such as median income, housing affordability, rehospitalization rates, food stamps usage, and
uninsured rates. Understanding these correlations is vital in comprehending 
the socio-economic and racial disparities prevalent in the studied region.

@@

These analyses reveals the presence of a drastic inequality in County 1(Fairfield County). Although Farfield County
is revealed to have the highest median income, greatest educational attainment, Farfield County also has the greatest 
percent population uninsured with health insurance and the greatest proportion of population whose renters pay rent 
50 percent of their income. According to a study by the Federal Reserve Bank of New York{ ???????}, the most highest paid 
workers in Fairfield County earn a salary that is nine times greater than their lower-wage counterparts. Technology causing the disparities
\\
II. Comparison by Year
\\
An analysis spanning four years indicates significant differences in median income, with the highest observed 
in 2020 and the lowest in 2017. While there doesn't seem to be a notable overall difference in the proportion 
of the population relying on food stamp programs or facing uninsured rates, distinctions emerge when considering 
specific timeframes. The use of food stamps was higher in 2020 and 2019 compared to 2017, showcasing a 
shift in reliance on such programs over the years. Similarly, the uninsured rates were notably higher in 2020 and 
2018 compared to 2019. These findings suggest despite an overall increase in income over time, there is a simultaneous 
growing demand for food stamp programs and a pressing need for improved access to affordable health insurance. 
This suggests that, while income is on the rise, significant portions of the population still face challenges in
 meeting basic needs and accessing essential services. Addressing these trends requires comprehensive strategies 
 that go beyond income growth alone.

>>>>

\section{Introduction}\label{sec:intro}

I. Prevalence of Social Determinants of Health

Current research focuses on how social determinants of health (SDoH) plays a  massive
impact on one's health; it is estimated that 80 percent of a population's health outcomes are 
dictated by SDoH \citep{HOOD2016129}. Often, SDoH, when referring to an individual, can result in racial 
disparities in care when looking at a population\citep{Monroe2023-uq}. It has been shown that major inefficiencies
in the health system are attributed to overlooked prevention opportunities and unequal access
to care.\citep{Allin2014-xn}

@@

Despite the growing body of research on SDoH, there is a notable gap concerning the specific impact of COVID-19 on 
these determinants within diverse communities. This research aims to bridge this gap by comprehensively assessing how 
the pandemic has influenced key social determinants of health across the 8 CT counties and a four year timeframe prior to 
the start of the COVID pandemic. 
By identifying the specific ways in which different communities were affected, this study contributes valuable insights 
for targeted interventions, policy-making, and the development of equitable healthcare strategies.
\@
II. Introduction to the Impact of Social Needs(specify for CT)
\\
This study will compare each of the following social needs by Year(2017-2020) and by 8 CT Counties: 
Median Income,
Education,
Food Access
Housing Stability,
Health Insurance,
Reshospitalization rates.

Income is a fundamental
social need that significantly influences
one's health. The level of income
directly affects an individual's access
to essential resources such as nutritious
food, healthcare, and secure housing.
People with lower incomes often face
barriers to healthcare services and
struggle to afford a balanced diet,
leading to higher risks of chronic
illnesses. Income disparities contribute
to inequalities in health outcomes, as
those with limited financial resources
may experience higher stress levels and
reduced access to preventive care. 
Data on income and mortality from the National Longitudinal Mortality Study reported by Backlund, Sorlie, and Johnson (1999) (4)
reveals that a continued drop in mortality results as income increases with a more significant drop towards the lower portion of the distribution(4).
Addressing income inequality is crucial
for promoting overall well-being and
reducing health disparities in communities.

Education is a key social determinant
of health, influencing various aspects
of an individual's life. Higher levels
of education are associated with better
health outcomes, as education enhances
one's ability to make informed decisions
about lifestyle choices, healthcare
utilization, and disease prevention. In CT, it is revealed there has been a spike in chronic absenteeism during the onset of the pandemic.
''Specifically, between the 2018-19 and 2021-22 school years, the share of students chronically absent grew by 13.5 percentage points.''(1)
There is also evidence of decreased Connecticut Performance Index (CPI) below the state target levels from 2019 to 2022 in mathematics, science,
and language arts.(1)
Limited educational opportunities can
result in lower health literacy, reducing
an individual's capacity to navigate
complex healthcare systems and understand
preventive measures. Educational
disparities contribute to a cycle of
poor health outcomes, as individuals
with lower levels of education may face
challenges in securing stable employment
and accessing adequate healthcare
resources. COVID-19 has only exaggerated
the pre-existing opportunity gap between urban schools and suburban schools in providing equal educational
opportunities despite both being in Connecticut.(2)
From (5), data reveals that  eath rates are lower for those with more education for all-cause mortality. 
The gap between high school graduates and those with some post-secondary education are larger compared to the latter group consisting
of a mixture of secondary education including college and advanced degrees.

Access to sufficient
food is a critical social need that
directly impacts health. Food insecurity,
often linked to low income, can lead to
malnutrition, obesity, and various health
issues. Lack of access to fresh and
healthy foods contributes to the prevalence
of chronic conditions such as diabetes
and cardiovascular diseases. Current
issues related to food include disparities
in food availability, affordability, and
nutritional quality, disproportionately
affecting marginalized communities.
Addressing these issues involves promoting
food equity, supporting local agriculture,
and implementing policies that ensure
everyone has access to nutritious food,
ultimately improving overall public health.

@@

In Figure 1
\ref{fig:Education by County Graph}
, clear distinctions among the eight counties are evident. 
Specifically, Fairfield County stands out with the highest proportion of individuals attaining a bachelor's 
degree as their highest level of education. Conversely, Windham County appears to lead in the percentage of 
residents with a high school diploma as their highest educational attainment. These findings align with the median income 
comparisons presented in Figure 2
\ref{fig:Social Needs by County Graph}
, covering the years 2017 to 2020. 
During this period, Fairfield County consistently displayed the highest median income, while Windham County 
consistently ranked lowest in terms of median income. This suggests a correlation between the counties with the 
highest median income and those with a higher prevalence of individuals achieving higher levels of educational attainment.

Table 4
\ref{fig:Social Needs by County Graph}
 reveals that there is a significant difference between 
social need variables by year (2017-2020) for median income, food, and uninsured rate. Table 4 also 
reveals that all five variables analyzes(median income, food, housing, rehospitalization rate, and 
uninsured rate) are significantly difference. Multiple Comparison tests are conducted to identify the 
specific similarities and differences between the 8 CT counties.
\\
I. Comparison by County
\\
The results of the Tukey multiple comparison analysis
\ref{tab: Multiple Comparisons}
 shed light on the relationship between county and key 
variables such as median income, housing affordability, rehospitalization rates, food stamps usage, and
uninsured rates. Understanding these correlations is vital in comprehending 
the socio-economic and racial disparities prevalent in the studied region.

@@

These analyses reveals the presence of a drastic inequality in County 1(Fairfield County). Although Farfield County
is revealed to have the highest median income, greatest educational attainment, Farfield County also has the greatest 
percent population uninsured with health insurance and the greatest proportion of population whose renters pay rent 
50 percent of their income. According to a study by the Federal Reserve Bank of New York{3}, the most highest paid 
workers in Fairfield County earn a salary that is nine times greater than their lower-wage counterparts. Technology causing the disparities
\\
II. Comparison by Year
\\
An analysis spanning four years indicates significant differences in median income, with the highest observed 
in 2020 and the lowest in 2017. While there doesn't seem to be a notable overall difference in the proportion 
of the population relying on food stamp programs or facing uninsured rates, distinctions emerge when considering 
specific timeframes. The use of food stamps was higher in 2020 and 2019 compared to 2017, showcasing a 
shift in reliance on such programs over the years. Similarly, the uninsured rates were notably higher in 2020 and 
2018 compared to 2019. These findings suggest despite an overall increase in income over time, there is a simultaneous 
growing demand for food stamp programs and a pressing need for improved access to affordable health insurance. 
This suggests that, while income is on the rise, significant portions of the population still face challenges in
 meeting basic needs and accessing essential services. Addressing these trends requires comprehensive strategies 
 that go beyond income growth alone.
